---
title: "Reproducible Research: Peer Assessment 1"
author: "Jerry Dumblauskas"
date: "February 16, 2016"
output: 
  html_document:
    keep_md: true
---


# Loading and preprocessing the data
##1 Code for reading in the dataset and/or processing the data
read the data:  please note the activity.csv file is assumed to be in the same directory

```{r}
qqq<-read.csv('activity.csv')
```


# What is mean total number of steps taken per day?
##2 Histogram of the total number of steps taken each day
Histogram of steps per day, using the aggregate function (step~day and summed)
```{r}
temp_data_sum<-aggregate(steps ~ date, qqq, sum)
hist(temp_data_sum$steps)

```

##3 Mean and median number of steps taken each day
Mean and median of steps per day
```{r}
my_mean<-mean(temp_data_sum$steps)
my_median<-median(temp_data_sum$steps)
print (my_mean)
print(my_median)
```


# What is the average daily activity pattern?
##4 Time series plot of the average number of steps taken
Time Series Plot with a mean

````{r}
temp_data_mean<-aggregate(steps ~ date, qqq, sum)
plot(temp_data_mean)
```

##5 The 5-minute interval that, on average, contains the maximum number of steps
Time Series Plot

````{r}
aggdata <-aggregate(qqq$steps, by=list(qqq$interval), FUN=sum, na.rm=TRUE)
```


# Imputing missing values
##6 Code to describe and show a strategy for imputing missing data
tbd

````{r}
```
##7 Histogram of the total number of steps taken each day after missing values are imputed
tbd

````{r}
```

# Are there differences in activity patterns between weekdays and weekends?
##8 Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
tbd

````{r}
```
