---
title: "Storm Data Analysis with respect to human and economic health"
author: "Jerry Dumblauskas"
date: "March 11, 2016"
output: html_document
---

#Synopsis 
We are going to analyze storm data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.  We want to ask 2 questions: 1) What storm events have the most harmful effects on human health, and 2) Across the United States, which types of events have the greatest economic consequences?

For this analysis, we are going to define human health as a combinations of injuries and fatalities (but won't give weight to one or the other).  We will also combine property and crop damage and aggregate the effects, in dollar terms, to show a summarized picture.

#Data processing
Let's load the data.  We will assume that the data file (which can be found on the NOAA's website), has already been downloaded and is in the same directory as this document.

```{r loaddata, cache=TRUE}
dta<-read.csv("repdata-data-StormData.csv.bz2")

```

Now, this is a huge dataset and we want to work with a smaller set.  We will just pull the variables we need.  We also for economic consequences need to fix and adjust the 'exponents' field, as it is in a non standard format. Here, we have to mutate the data to account for the 'exp' column of the data.  As best as can be figured, numbers like 2, mean 'multiply the previous column by 10`.  Items like 'b' and 'B' mean 'Billions'.  We need to do this for both property damage and crop damage
```{r process_and_condense_data, cache=TRUE}
danger_to_human_life<-subset(dta,select = c('EVTYPE', 'FATALITIES', 'INJURIES'))


economic_damage<-subset(dta,select = c('EVTYPE', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP'))

unique(dta$PROPDMGEXP)
unique(dta$CROPDMGEXP)

economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='']<-1
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='-']<-0
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='?']<-0
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='+']<-0
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='0']<-1
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='1']<-1
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='2']<-10
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='3']<-100
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='4']<-1000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='5']<-10000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='6']<-100000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='7']<-1000000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='8']<-10000000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='B']<-1000000000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='h']<-100
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='H']<-100
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='K']<-1000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='m']<-1000000
economic_damage$PROPDMGNUM[economic_damage$PROPDMGEXP=='M']<-1000000

economic_damage$PROPDMGTOTAL<-economic_damage$PROPDMGNUM * economic_damage$PROPDMG
  
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='']<-1
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='?']<-0
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='0']<-1
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='2']<-10
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='B']<-1000000000
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='k']<-1000
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='K']<-1000
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='m']<-1000000
economic_damage$CROPDMGNUM[economic_damage$CROPDMGEXP=='M']<-1000000

economic_damage$CROPDMGTOTAL<-economic_damage$CROPDMGNUM * economic_damage$CROPDMG
```

#Results
###Effects on Human life
First, lets handle the highest events that harmful effects on human life.  We will highlight the top 20.  So, let's aggegrate, sort, then plot the data.
```{r plotlife, cache=TRUE}
aggdata_hl<-aggregate(danger_to_human_life[,2:3], by=list(danger_to_human_life$EVTYPE), FUN = sum, na.rm=TRUE)
aggdata_hl$totals<-rowSums(aggdata_hl[,2:3])
sorted_hl<-aggdata_hl[order(-aggdata_hl$totals), ]
plot_data_hl<-head(sorted_hl,20)
plot_data_hl$totals<-plot_data_hl$totals/1000
barplot(plot_data_hl$totals, ylim=c(0, 100), names.arg=plot_data_hl$Group.1, las=2, ylab = "Number of incidents (in Thousands)", main = "Highest 20 events that have the most harmful effects on human life ")

```

###Econonic Consequences

We can plot the data, we again will get the top 20 economic events (in Billions of Dollars), after we aggregate and sort
```{r plot_econ_data, cache=TRUE}
# upper this
economic_damage$EVTYPE<-toupper(economic_damage$EVTYPE)
# sort and combine the data into a totals collumn
aggdata_econ_consequences<-aggregate(economic_damage[,c(7,9)], by=list(economic_damage$EVTYPE), FUN = sum, na.rm=TRUE)
aggdata_econ_consequences$totals<-rowSums(aggdata_econ_consequences[,2:3])
sorted_ec<-aggdata_econ_consequences[order(-aggdata_econ_consequences$totals), ]
#set to billions
sorted_ec$totals<-sorted_ec$totals/1000000000
plot_data_ec<-head(sorted_ec,20)
barplot(plot_data_ec$totals, ylim=c(0, 200), names.arg=plot_data_ec$Group.1, las=2, ylab = "Dollar Damage (in Billions of Dollars)", main = "Highest 20 events that have the most harmful effects on the economy ")

```

#Summary
Tornados seem to hurt people the most and Floods seem to cause the most dollar denominated damage!